# Example of `make` for data analysis

This repository shows how to use `make` for a data analysis project.

The paper (PDF) is generated from a LaTeX source file, a table, and a figure.

`make` uses a set of instructions in `makefile` to generate the table and figure using R code and a datafile, and then generate the PDF from the manuscript.

`make` is clever because it deconstructs each part of the analysis so that only parts that have changed need to be rerun. If the data change, everything is rerun. If figure-generating code changes, only that code and the manuscript are rerun. If only the manuscript changes, only `pdflatex` is rerun. It's smart like that.

Basically it works on a directed acyclic graph (DAG) model, sort of represented by this figure:

```{r setup, include=FALSE}
library("svglite")
knitr::opts_chunk$set(
  dev = "svglite",
  fig.path="fig-",
  fig.ext = "svg"
)
```

```{r dag, echo = FALSE}
requireNamespace("ggdag", quietly = TRUE)
requireNamespace("dagitty", quietly = TRUE)

parse_makefile <-
function(
  file = "makefile",
  ...
) {
    con <- file(file, open = "rb")
    on.exit(close(con))
    m <- character()
    # handle continuation characters
    while (length(con)) {
        this_line <- readLines(con, n = 1L)
        if (!length(this_line)) {
            break
        }
        m[length(m) + 1] <- this_line
        while (grepl("\\\\$", m[length(m)]) && length(con)) {
            m[length(m)] <- sub("\\\\$", " ", m[length(m)])
            m[length(m)] <- paste0(m[length(m)], readLines(con, n = 1L))
        }
    }
    return(m)
}

makefile_to_dag <-
function(
  lines = parse_makefile(),
  exclude = NULL,
  ...
) {
    rules <- lines[grepl("^[A-Za-z0-9./]+ ?[:]", lines)]
    rules_list <- strsplit(rules, " ?: ?")
    formulae <- lapply(rules_list, function(x) {
        # cleanup target
        target <- x[1L]
        if (target %in% exclude) {
            return(NULL)
        }
        target <- gsub("/", "_", target)
        
        # cleanup prereqs
        prereqs <- strsplit(x[2L], "[ \t]+")[[1L]]
        prereqs <- prereqs[!is.na(prereqs)]
        prereqs <- gsub("/", "_", prereqs)
        
        # return
        if (!length(prereqs)) {
            return(NULL)
        }
        return(paste0(prereqs, " -> ", target, ";"))
    })
    paste0("dag{", paste0(unlist(formulae[lengths(formulae) > 0]), collapse = ""), "}")
}

plot_makefile_dag <-
function(
  file = "makefile",
  exclude = NULL,
  layout = "gem",
  size = 4,
  ...
) {
    mf <- makefile_to_dag(parse_makefile(file), exclude = exclude)
    dag <- dagitty::dagitty(mf)
    ggdag::ggdag_classic(dag, layout = layout, size = size, ...) + ggdag::theme_dag_blank()
}

plot_makefile_dag(exclude = "all")
```

(Note: directory `/` symbols are converted to `_` in the above.)

The R file `analysis.R` shows what is going on in `makefile` using possibly more familiar R syntax.

Zach Jones has [a good tutorial](http://zmjones.com/make/) about all of this.
